\bf On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2024-11-20 Manuscript/manuscript.tex
just started, still reading Data section

The disqualification of Allen was widely discussed in online 
communities, with some questioning the accuracy of the timing 
equipment. One of the most active platforms was \url{www.LetsRun.com}, 
a website that functions as both a message board and a news outlet. 
The message board, similar to Reddit, is particularly active during 
major running events, including the Olympics and the World Track 
and Field Championships. The news section of the website featured 
articles such as \citep{johnson2022data} and \citep{johnson2022was} 
by Robert Johnson, the site’s founder. Johnson argued that there may 
have been an error with the timing equipment at the 2022 Championships. 
He cited graphs and descriptive statistics, like median reaction 
times, comparing the reaction times at the 2022 Championships to 
those from other competitions, including the U.S. Championships and 
previous World Championships. However, these comparisons lacked formal 
statistical analysis.

@@

This paper has two primary objectives. First, we investigate whether 
reaction times at the 2022 World Championships were significantly 
different from other competitions, focusing on athletes who competed 
in multiple events. Using a matched-pairs design, we compare reaction 
times across the 2022, 2019, and 2023 World Championships, as well as 
the 2022 national-level competitions. This approach isolates the 
effect of the competition year while controlling for individual 
performance. The second objective is to assess the 
appropriateness of the 0.1-second threshold by modeling reaction 
times from World Championships held from 1999 onward. We use a gamma 
linear mixed-effects model with random effects for both venue and heat. 
This model estimates the probability of reaction times falling below 
the threshold, helping evaluate the consistency of reaction times and 
whether the current threshold is statistically justified.

The rest of this paper is organized as follows. Section~\ref{sec:Data} 
describes the data collection and datasets used for the two
objectives. Section~\ref{sec:rank} outlines the methods for
objective~1, analyzing differences in reaction times across
competitions, including the were collected. Section~\ref{sec:gamlss}
details the methods for objective~2, the generalized additive model
for location, scale, and shape (GAMLSS) to evaluate the appropriatenes
of the 0.1-second threshold. Section~\ref{sec:Results} presents the
results of both objectives and discusses the implications.
Finally, Section~\ref{sec:concludingremarks} highlights the 
paper’s impact, limitations, and potential for future research.
All data and code for our analysis 
are provided in the supplementary materials.

@@

There are two types of data that are explored throughout this paper. For the
rank-sum tests described in Section~\ref{sec:rank}, the data comes from male 
athletes who competed in the 110 meter hurdles and 100 meter dash, and female
athletes who competed 100 meter hurdles and 100 meter dash.
Athletes must have competed in the 2022 World Championships and another competition 
(2019 World Championships, 2023 World Championships, or 2022 national 
championships). For the GLMM described in Section~\ref{sec:glmm}, data comes
from every world championships from 1999 to 2023 and includes reaction times from
the 110 meter hurdles and 100 meter dash. This investigation started 
shortly after the 2022 World Championships, but as the 2023 data has now become 
available, we are pleased to include it in both of our analyses, and it did not
significantly change our results \citep{WAData}.

@@

\subsubsection{2022 National Competitions}\label{sec:datanational}
From June 23-26, 2022, the United States held its Track and Field Championships 
at Hayward Field in Eugene, Oregon to decide who to send to the World 
Championships being held at the same venue in August. Thus, we can establish a 
baseline for the four United States athletes who competed in at least one round 
of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen.  
Every athlete reacted faster in all the World Track and Field Championships 
races compared to the United States Track and Field Championships races. The 
dataset of the four United States athletes is too small to perform any 
meaningful statistical analysis, but the data was expanded to include reaction 
times from more athletes who competed at other national competitions (many of which were 
held from May-July of 2022). This data was difficult to find, as the data is not
centrally located but instead found on different national websites (often 
recorded in the national language without translation).

Each athlete was designated as a ``cluster'' and the stage of competition (national
or international) was specified for each competition within the cluster (athlete).
Thus, this is subunit level grouping where we assign the treatment, national or
international, within each cluster. Cluster sizes range from three to six, with
one cluster of size three, six clusters of size four, two clusters of size five,
and two clusters of size six.
\of{This explanation is going to need to be re-worked now that we have more data.
Should we make a table of some sort? It's going to be very wordy without one}

We chose to include preliminary rounds in this section as well as in
Section~\ref{sec:data2019}, as removing them would have resulted in a much
narrower analysis with fewer athletes and clusters.

\subsubsection{Other World Track and Field Championships}\label{sec:data2019}
The 2019 and 2023 World Track and Field Championships data is included in the 
larger dataset described in Section~\ref{sec:dataworld}, but we can also use it
similarly to how we used national level in Section~\ref{sec:datanational}. We
can compare the times of athletes who participated in the 2022 Championships with
their times at another World Championships: 2019 and/or 2023.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics{RankScatterPlots}
  \caption{This collection of scatterplots of data used in the rank comparison
  section shows that times in 2022 were on average lower. On the horizontal
  axis below each graph "H", "S", and "F" refer to the heats, semifinals, and 
  finals respectively. Please note that in the last row the 2022 times are to
  the right of the 2023 times.}
  \label{fig:RankScatterplots}
\end{figure}

Figure~\ref{fig:RankScatterplots} shows the data that will be analyzed
with the rank-based methods (described in Section~\ref{sec:rank} with results in 
Section~\ref{subsec:Results_Rank}). In each of the six graphs, the reaction 
times for athletes who competed at the 2022 World Championships
are compared to their reaction times at another at which they competed. 
It is worth noting that each instance of the 2022 data is different, because
those who competed at both the 2019 and 2022 World Championships will be
different athletes than those who competed at both the 2022 and 2023 World
Championships. This data was only for athletes who competed
in at least one race at each championship, however it is possible for athletes
to have competed in one race in 2019 and two races in 2022. The 2022 World
Championship heats and semifinals both had on average lower reaction times than
their respective other (2019, 2023 or 2022 national) races, and the 2022 World 
Championship finals were substantially faster.  It
is also worth noting is that the fastest reaction time in the 2022 World
Championships Finals and Semifinals both came from Devon Allen, but the 
difference between 0.101 and 0.099
seconds was the difference in what caused him to be disqualified.

>>>>

The disqualification of Allen was widely discussed in online 
communities, with some questioning the accuracy of the timing 
equipment. One of the most active platforms was \url{www.LetsRun.com}, 
a website that functions as both a message board and a news outlet. 
The message board, similar to Reddit, is particularly active during 
major running events, including the Olympics and the World Track 
and Field Championships. The news section of the website featured 
articles such as \citet{johnson2022data} and \citet{johnson2022was} 
by Robert Johnson, the site’s founder. Johnson argued that there may 
have been an error with the timing equipment at the 2022 Championships. 
He cited graphs and descriptive statistics, like median reaction 
times, comparing the reaction times at the 2022 Championships to 
those from other competitions, including the U.S. Championships and 
previous World Championships. However, these comparisons lacked formal 
statistical analysis.

@@

This paper has two primary objectives. First, we investigate whether 
reaction times at the 2022 World Championships were significantly 
different from other competitions, focusing on athletes who competed 
in multiple events. Using a matched-pairs design, we compare reaction 
times across the 2022, 2019, and 2023 World Championships, as well as 
the 2022 national-level competitions. This approach isolates the 
effect of the competition year while controlling for individual 
performance. The second objective is to assess the 
appropriateness of the 0.1-second threshold by modeling reaction 
times from World Championships held from 1999 onward. We use a gamma 
linear mixed-effects model with random effects for both venue and heat. 
\eds{edit previous as needed for new GAMLSS analysis}
This model estimates the probability of reaction times falling below 
the threshold, helping evaluate the consistency of reaction times and 
whether the current threshold is statistically justified.

The rest of this paper is organized as follows. Section~\ref{sec:Data} 
describes the data collection and datasets used for the two
objectives. Section~\ref{sec:rank} outlines the methods for
objective~1, analyzing differences in reaction times across
competitions, including the were collected. 
\eds{end of previous sentence does not make sense; some text may have been 
mistakenly deleted?}
Section~\ref{sec:gamlss}
details the methods for objective~2, the generalized additive model
for location, scale, and shape (GAMLSS) to evaluate the appropriatenes
of the 0.1-second threshold. Section~\ref{sec:Results} presents the
results of both objectives and discusses the implications.
Finally, Section~\ref{sec:concludingremarks} highlights the 
paper’s impact, limitations, and potential for future research.
All data and code for our analysis 
are provided in the supplementary materials.

@@

There are two types of data that are explored throughout this paper. For the
rank-based comparisons described in Section~\ref{sec:rank}, the data described in 
Section~\ref{sec:databeyond} come from male 
athletes who competed in the 110 meter hurdles and \eds{or?} 100 meter dash, and 
female athletes who competed 100 meter hurdles and \eds{or?} 100 meter dash.
Athletes must have competed in the 2022 World Championships and another competition 
(2022 national championships, 2019 World Championships, or 2023 World Championships) 
in order to be included in the rank-based comparisons. 
For the GAMLSS analysis described in Section~\ref{sec:gamlss}, the data described 
in Section~\ref{sec:dataworld} come
from every world championships from 1999 to 2023 and includes reaction times from
the 110 meter hurdles and 100 meter dash. This investigation started 
shortly after the 2022 World Championships, but as the 2023 data has now become 
available, we are pleased to include it in both of our analyses, and it did not
significantly change our results \citep{WAData}. \eds{Should we remove this 
last sentence, since now likely 2024 data is available?}

@@

\subsubsection{2022 National Competitions}\label{sec:datanational}
From June 23-26, 2022, the United States held its Track and Field Championships 
at Hayward Field in Eugene, Oregon to decide who to send to the World 
Championships being held at the same venue in August. Thus, we can establish a 
baseline for the four United States athletes who competed in at least one round 
of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen.  
Every athlete reacted faster in all the World Track and Field Championships 
races compared to the United States Track and Field Championships races. The 
dataset of the four United States athletes is too small to perform any 
meaningful statistical analysis, so the data for the rank-based analyses was 
expanded to include reaction 
times from more athletes who competed at other national competitions (many of which were 
held from May-July of 2022). This data was difficult to find, as the data is not
centrally located but instead found on different national websites (often 
recorded in the national language without translation).

Each athlete was designated as a ``cluster'' and the stage of competition (national
or international) was specified for each competition within the cluster (athlete).
Thus, this is subunit level grouping where we assign the treatment, national or
international, within each cluster. Cluster sizes range from three to six, with
one cluster of size three, six clusters of size four, two clusters of size five,
and two clusters of size six.
\of{This explanation is going to need to be re-worked now that we have more data.
Should we make a table of some sort? It's going to be very wordy without one}
\eds{We are only doing two analyses with this data, right, one for women 
and one for men?  So perhaps only two sentences needed here? But then 4 sentences
in the next subsection?  Maybe you can just report ranges of cluster sizes and 
median cluster size?}

We chose to include reaction times from preliminary heats, in addition to 
reaction times from semifinals and finals, in this section and in
Section~\ref{sec:data2019}, as removing them would have resulted in a much
narrower analysis with fewer athletes and clusters.

\subsubsection{Other World Track and Field Championships}\label{sec:data2019}
The 2019 and 2023 World Track and Field Championships data is included in the 
larger dataset described in Section~\ref{sec:dataworld}, but we can also use it
similarly to how we used national competitions in Section~\ref{sec:datanational}. 
We can compare the reaction times of athletes who participated in the 2022 World 
Championships with their times at another World Championships: 2019 and/or 2023.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics{RankScatterPlots}
  \caption{Reaction times for athletes who competed at the 2022 World Championships
	and at another championship (2022 national, 2019 World, or 2023 World) at which 
	they competed. On the horizontal
  axis below each graph "H", "S", and "F" refer to the heats, semifinals, and 
  finals respectively. Please note that in the last row the 2022 times are to
  the right of the 2023 times.
	\eds{Please check my edited caption; also, I suggest rearranging the rows to 
	use the same ordering as the data was described: put national vs international 
	in first row, 2019 vs 2022 in second row, and 2022 vs 2023 in third row}
	}
  \label{fig:RankScatterplots}
\end{figure}

Figure~\ref{fig:RankScatterplots} shows the data that will be analyzed
with the rank-based methods (described in Section~\ref{sec:rank} with results in 
Section~\ref{subsec:Results_Rank}). In each of the six graphs, the reaction 
times for athletes who competed at the 2022 World Championships
are compared to their reaction times at another at which they competed. 
It is worth noting that for each row of the figure, the athletes compromising 
the 2022 data is different.  This is because the set of athletes who competed at 
both the 2019 and 2022 World Championships is not the same as the set of athletes 
who competed at both the 2022 and 2023 World Championships. %This data was only for athletes who competed
%in at least one race at each championship, however it is possible for athletes
%to have competed in one race in 2019 and two races in 2022. 
Athletes must have competed in at least one race at each championship to be 
included.  The 2022 World
Championship heats and semifinals both had on average lower reaction times than
their respective other (2019, 2023 or 2022 national) races, and the reaction times 
from the 2022 World Championship finals were substantially faster. 
\eds{I would consider removing or rephrasing the last part of the previous 
sentence. With the women's data, it is not as clear the 2022F is lower than 2022S.}
\eds{I also think we need to comment on the men and women's data. But within row
of the figure, don't the distributions appear similar?  If that is so, would a 
reviewer ask why we didn't pool the data for men and women?}
It is also worth noting is that the fastest reaction time in the 2022 World
Championships Finals and Semifinals both came from Devon Allen, but the 
difference between 0.101 and 0.099
seconds was the difference in what caused him to be disqualified.
